Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-05 ms.tex
clearing up

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars
\citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI and CTL targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-sky analysis we will focus on F and G stars since the 27.5 day
baseline of \TESS\ fields will limit measureable rotation periods to less than
around 15 days.
With the April 2018 data release from Gaia (DR2), we estimate that we will be
able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
kpc.
To characterize the rotation period bimodality as a function of galactic
position, we propose to analyse the light curves of 10,000 FFI and two-minute
targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
selecting FGKM dwarfs with a range of galactic positions.
The FFI targets will be essential in order to reach $\sim$3kpc distances.
Of these, the stars lying in the CVZs will be our highest priority.
M dwarfs rotation periods in the CVZz will allow us to study the rotation
period bimodality at lower masses, since the bimodality appears at longer
rotation periods for lower mass stars.

@@

Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods for both FFI and
two-minute cadence stars, with the maximum measurable period strongly
depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS\ /\HERMES\ survey, from which it
will be possible to calculate isochronal ages, enabling further calibration of
the gyrochronology relations.
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars in the TESS CTL will have proper motions, parallaxes, positions and
radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
% PI Angus is currently using the \kepler\ rotation periods of stars in the
% first \Gaia\ data release to calibrate the relations between rotation period
% and vertical action dispersion, two tracers of age.
% Since this relation is likely vary with both galactic longitude and latitude,
% the rotation periods produced from \TESS\ light curves will enable a
% comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
In order to calibrate gyrochronology using rotation periods of FGK dwarfs with
\tess\ / \hermes\ spectra we propose to prioritize rotation periods of stars
in the CVZs with spectra, followed by those without.

>>>>

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars
\citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI and CTL targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-sky analysis we will focus on F and G stars since the 27.5 day
baseline of \TESS\ fields will limit measureable rotation periods to less than
around 15 days.
With the April 2018 data release from Gaia (DR2), we estimate that we will be
able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
kpc.
To characterize the rotation period bimodality as a function of galactic
position, we propose to analyse the light curves of 10,000 FFI and two-minute
targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
selecting FGKM dwarfs with a range of galactic positions.
The FFI targets will be essential in order to reach $\sim$3kpc distances.
% Of these, the stars lying in the CVZs will be our highest priority.
M dwarfs rotation periods in the CVZz will allow us to study the rotation
period bimodality at lower masses, since the bimodality appears at longer
rotation periods for lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky coverage
necessary to map the period bimodality across the galaxy.

@@

Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods for both FFI and
two-minute cadence stars, with the maximum measurable period strongly
depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which it
will be possible to calculate isochronal ages, enabling further calibration of
the gyrochronology relations.
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
% PI Angus is currently using the \kepler\ rotation periods of stars in the
% first \Gaia\ data release to calibrate the relations between rotation period
% and vertical action dispersion, two tracers of age.
% Since this relation is likely vary with both galactic longitude and latitude,
% the rotation periods produced from \TESS\ light curves will enable a
% comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
% In order to calibrate gyrochronology using rotation periods of FGK dwarfs with
% \tess--\hermes\ spectra we propose to prioritize rotation periods of stars
% in the CVZs with spectra, followed by those without.
